### 2019-02-22 concepts.tex
more tstl intro

Vertical nondeterminism, in fact, is potentially \emph{completely} unrelated
to flakiness, in that some bugs (improper handling of hardware errors
in a file system, for example) produce \emph{reliably}
``nondeterministic'' (from a software library users' point of view)
behavior.  Vertical nondeterminism may rely on ``real'' nondeterminism
that could produce a flaky test, but it does not have to do so.

\subsection{Formal Definitions}

>>>>

Vertical nondeterminism, in fact, is potentially \emph{completely} unrelated
to flakiness, in that some bugs (improper handling of hardware errors
in a file system, for example) produce \emph{reliably}
``nondeterministic'' (from a software library users' point of view)
behavior.  Vertical nondeterminism may rely on ``real'' nondeterminism
that could produce a flaky test, but it does not have to do so.  In
other words, vertical nondeterminism is often not a problem for
testing (flakiness), but simply a bug in the behavior of a software
system, where an implicit specification on failed operations is
consistently violated under certain conditions.  The behavior is
``nondeterministic'' from a caller's point of view, however, in that
there is an expectation that a call will always ``do the same thing,
under the same circumstances'' which is the \emph{conceptual} common
factor in all kinds of expected determinism.

\subsection{Formal Definitions}
\label{sec:formal}

### 2019-02-22 implementation.tex
more tstl intro

We implemented our approach in the TSTL \cite{NFM15,ISSTA15,tstlsttt}
system, an open-source
language and tool for property-based testing
\cite{Hypothesis,ClaessenH00} of Python code.   TSTL, including full
support for our nondeterminism detection methods, is available at
\url{https://github.com/agroce/tstl}, and the {\tt examples} directory
contains the source code needed to duplicate all of our experiments;
TSTL can also be installed using Python's {\tt pip} tool, by typing
{\tt pip install tstl}; you will need to clone the github repository
to have access to the case studies used in the paper, however.

@@

@@

We therefore instead made horizontal nondeterminism detection a first-class
property in TSTL, using TSTL's existing notation for marking some
types of values as \emph{opaque} (not usefully compared for equality),
used in automatic abstraction-based testing.  In TSTL, a test is a
sequence of \emph{actions}; parameter values to method and function
calls of the Software Under Test (SUT) are stored in finite-sized variable \emph{pools}.  A
TSTL test is (essentially) a sequence of assignments to pool values,
and method/function calls using pool values (including method calls on
pool objects).  This formalism is a widely used and efficient way to
represent unit tests \cite{Pacheco,AndrewsTR}.  In TSTL, \emph{visible
value determinism} is based on comparing the values of \emph{all pool
variables} after each test action (pool values other than the one
assigned to must also be compared, since a common source of
nondeterminism is when a call results in a change to a value passed as
a parameter, changes a shared reference held by two values).
TSTL omits comparison only of pools marked as {\tt OPAQUE} in the
TSTL language test harness (and values that do not support equality
checking).  \emph{Final State Determinism} simply performs the same
comparison, but only on the final values of all pool variables (or a
set of designated pools) after a test has
finished executing.

>>>>

We implemented our approach in the TSTL \cite{NFM15,ISSTA15,tstlsttt}
system, an open-source
language and tool for property-based testing
\cite{Hypothesis,ClaessenH00} of Python code.   TSTL, including full
support for our nondeterminism detection methods, is available at
\url{https://github.com/agroce/tstl}, and the {\tt examples} directory
contains the source code needed to duplicate all of our
experiments\footnote{In some cases, you will also need to install an
  appropriate version of the tested software, by doing, e.g., {\tt pip
    install pyfakefs}.};
TSTL can also be installed using Python's {\tt pip} tool, by typing
{\tt pip install tstl}; you will need to clone the github repository
to have access to the case studies used in the paper, however.

@@

\begin{figure}
\begin{code}
pool: <int> 5
pool: <l> 5
pool: <s> 5
\vspace{0.1in}
<int> := <[1..20]>
<l> := []
<l> := list(<s>)
<s> := set()
<s> := set(<l>)
\vspace{0.1in}
<l>.append(<int>)
\{ValueError\} <l>.remove(<int>)
<s>.add(<int>)
\{KeyError\} <s>.remove(<int>)
\vspace{0.1in}
property: len(<l>) >= len(set(<l,1>))
property: len(<s>) == len(list(<s,1>))
\end{code}
\caption{Simple TSTL harness testing lists and sets}
\label{fig:simple}
\end{figure}

In TSTL, a test consists of a sequence of
\emph{actions}, where actions execute arbitrary Python code, but are
expected to work by modifying the \emph{state} of a test, which is
stored in a fixed set of \emph{pools} containing Python objects.  The
TSTL actions defined by a test harness correspond to the set $A$ of
actions in Section \ref{sec:formal}, and transitions between states of
result from carrying out the Python code in an action.  Figure
\ref{fig:simple} shows a very simple TSTL harness that tests Python's
built-in {\tt set} and {\tt list} types.  The harness defines the
(visible) state of a test as consisting of three named pools of Python
objects, {\tt int}, {\tt l}, and {\tt s}.  These pools have no
inherent associated type, though a type can be annotated as an
additional constraint that TSTL will check holds during test
execution.  Initially all the pools hold the special Python value {\tt
  None} indicating they have not been initialized.  In terms of our formalism,
$I = \{${\tt [<int> = [None, None, None, None, None]}, {\tt <l> =
  [None, None, None, None, None]}, {\tt <s> = [None, None, None, None,
  None]]}$\}$.  Actions with the {\tt :=} assignment operator
initialize pools.  For example, the line of code {\tt l := []} defines a set of 5
actions, each of which initializes a different member of the list pool
{\tt l} to an empty list --- {\tt l0 = []} ... {\tt l4 = []}.  Actions
without a {\tt :=} require that pools used in them already be
initialized.  TSTL provides constructs like {\tt <[1..20]>} for
choosing values in a range for use in a test, and allows a user to
specify which exceptions an action may raise without causing the test
to fail (as in {\tt \{KeyError\} <s>.remove(<int>)}).

Compiling this harness using
the TSTL compiler produces a Python module that allows test
generation.  If we use the random testing tool provided by TSTL, {\tt
  tstl\_rt}, it will randomly generate valid sequences of actions, and
report no problems.  If we change the first property to read {\tt
  property: len(<l>) == len(set(<l,1>))}, however, TSTL will report a
failing test, shown in Figure \ref{fig:setlistfail}.

\begin{figure}
{\scriptsize
\begin{code}
int0 = 1                                                                 \# STEP 0
ACTION: int0 = 1 
int0 = None : <type 'NoneType'>
=> int0 = 1 : <type 'int'>
==================================================
l0 = []                                                                  \# STEP 1
ACTION: l0 = [] 
l0 = None : <type 'NoneType'>
=> l0 = [] : <type 'list'>
==================================================
l0.append(int0)                                                          \# STEP 2
ACTION: l0.append(int0) 
int0 = 1 : <type 'int'>
l0 = [] : <type 'list'>
=> l0 = [1] : <type 'list'>
==================================================
l0.append(int0)                                                          \# STEP 3
ACTION: l0.append(int0) 
int0 = 1 : <type 'int'>
l0 = [1] : <type 'list'>
=> l0 = [1, 1] : <type 'list'>
==================================================
ERROR: (<type 'exceptions.AssertionError'>,
  AssertionError(), <traceback object at 0x1032b83f8>)
TRACEBACK:
  File "/Users/adg326/scratch/pyexample/sut.py",
  line 13695, in check
    assert len(self.p\_l[0]) == len(set(self.p\_l[0]))
\end{code}
}
\caption{A failing test (showing that converting a list to a set may
  not preserve the size of the object).}
\label{fig:setlistfail}
\end{figure}

@@

We therefore instead made horizontal nondeterminism detection a first-class
property in TSTL, using TSTL's existing notation for marking some
types of values as \emph{opaque} (not usefully compared for equality),
used in automatic abstraction-based testing.  Recall that in TSTL, a
test is (essentially) a sequence of assignments to pool values,
and method/function calls using pool values (including method calls on
pool objects).  This formalism is a widely used and efficient way to
represent unit tests \cite{Pacheco,AndrewsTR}.  In TSTL, \emph{visible
value determinism} is based on comparing the values of \emph{all pool
variables} after each test action (pool values other than the one
assigned to must also be compared, since a common source of
nondeterminism is when a call results in a change to a value passed as
a parameter, changes a shared reference held by two values).  By
default $V$ as defined in Section \ref{sec:formal} is just the set of all possible pool values, and $v$
simply extracts the pool values from $S$.
TSTL allows restriction of which state is visible by removing pools marked as {\tt OPAQUE} in the
TSTL language test harness (and values that do not support equality
checking) from $v(S)$.  \emph{Final State Determinism} simply performs the same
comparison, but only on the final values of all pool variables (or a
set of designated pools) after a test has
finished executing.
